---
title: "Getting Started with R"
subtitle: "Execute Simple Commands"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  revealjs::revealjs_presentation:
    theme: simple
    highlight: default
    transition: slide
    center: true
    css: modern-reveal-css.css
    self_contained: false
    lib_dir: libs
    reveal_plugins: ["notes", "search", "zoom"]
    reveal_options:
      slideNumber: true
      previewLinks: false
      hash: true
      height: 900
      width: 1600
      margin: 0.1
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width = 10, fig.height = 6, fig.retina = 3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

# Getting Started with R {.title-slide data-background-gradient="linear-gradient(135deg, #3a0ca3, #4361ee)"}

## Execute Simple Commands

### For New R Users

---

## R as a Calculator

R can be used as a powerful calculator. Type an expression and hit Enter.

```{r}
# Addition
5 + 3

# Subtraction
10 - 7

# Multiplication
4 * 6

# Division
15 / 3

# Exponents (Power)
2^3  # 2 raised to the power of 3
```

---

## Order of Operations

R follows the standard mathematical order of operations (PEMDAS):
- **P**arentheses
- **E**xponents
- **M**ultiplication and **D**ivision (from left to right)
- **A**ddition and **S**ubtraction (from left to right)

```{r}
# Without parentheses
3 + 4 * 2

# With parentheses to change order
(3 + 4) * 2
```

---

## More Mathematical Operations

```{r}
# Square root
sqrt(16)

# Absolute value
abs(-42)

# Logarithms
log(10)        # Natural logarithm (base e)
log10(100)     # Base 10 logarithm
log(8, base=2) # Base 2 logarithm

# Trigonometric functions (in radians)
sin(pi/2)
cos(pi)
tan(pi/4)
```

---

## Variables

Variables let you store values for later use. Use the assignment operator `<-` to create variables.

```{r}
# Assign a value to a variable
x <- 10
y <- 5

# Use variables in calculations
x + y
x * y
x / y

# You can also use the = operator for assignment, but <- is preferred
z = 15
z
```

<div class="tip">
<strong>Tip:</strong> Use <code>&lt;-</code> for assignment and <code>=</code> for specifying function arguments.
</div>

---

## Variable Naming Rules

Good variable names make your code more readable and easier to understand.

### Rules for variable names:
- Must start with a letter
- Can contain letters, numbers, underscore (_), and period (.)
- Cannot contain spaces or special characters
- Are case-sensitive (age ≠ Age)

```{r}
# Valid variable names
my_variable <- 10
myVariable <- 20
my.variable <- 30
variable1 <- 40

# Print a variable's value
my_variable
```

---

## Variable Naming Conventions

While many naming styles are valid, consistency is key. Choose one style and stick with it.

```{r}
# Common naming styles:

# snake_case (recommended for R)
first_name <- "John"

# camelCase
firstName <- "John"

# PascalCase 
FirstName <- "John"

# dot.case (less common in modern R)
first.name <- "John"
```

<div class="note">
<strong>Note:</strong> The tidyverse style guide recommends snake_case for variable and function names.
</div>

---

## Data Types: Numeric

R has several data types for representing different kinds of values.

### Numeric (includes both integers and decimal numbers)

```{r}
# Integers and doubles are both numeric
a <- 10    # This is actually stored as a double
b <- 10.5

# Check the type
class(a)
class(b)

# Force an integer with L suffix
c <- 10L
class(c)
```

---

## Data Types: Character

### Character (for text/string data)

```{r}
# Create character strings with quotes
name <- "Alice"
greeting <- 'Hello, world!'

# Check the type
class(name)

# Combining strings
paste("Hello", "there")
paste0("R", "Studio") # paste0 doesn't add spaces

# String length
nchar("Programming")
```

---

## Data Types: Logical

### Logical (Boolean values: TRUE or FALSE)

```{r}
# Create logical values
is_correct <- TRUE
has_errors <- FALSE

# Check the type
class(is_correct)

# Logical operations
TRUE & TRUE   # AND: both must be TRUE
TRUE | FALSE  # OR: at least one must be TRUE
!TRUE         # NOT: inverts the value
```

---

## Comparisons

Comparison operators return logical values (TRUE or FALSE).

```{r}
# Equal to
5 == 5

# Not equal to
5 != 3

# Greater than
10 > 5

# Less than
3 < 7

# Greater than or equal to
5 >= 5

# Less than or equal to
4 <= 10
```

---

## Basic Data Structures: Vectors

Vectors are one-dimensional collections of values of the same type.

```{r}
# Create a vector with c()
numbers <- c(2, 4, 6, 8, 10)
fruits <- c("apple", "banana", "cherry")
logicals <- c(TRUE, FALSE, TRUE, TRUE)

# Check the type
class(numbers)
class(fruits)

# Get the length
length(numbers)

# Access elements by position (R uses 1-based indexing)
numbers[1]      # First element
fruits[2]       # Second element
numbers[c(1,3)] # First and third elements
```

---

## Operating on Vectors

Operations on vectors are applied to each element.

```{r}
# Create a vector
vec <- c(10, 20, 30, 40, 50)

# Add a value to each element
vec + 5

# Multiply each element
vec * 2

# Operations between vectors (element-wise)
vec1 <- c(1, 2, 3)
vec2 <- c(10, 20, 30)
vec1 + vec2
vec1 * vec2
```

---

## Sequences and Repetition

### Creating sequences of numbers:

```{r}
# Using the colon operator for consecutive integers
1:10

# Using seq() for more control
seq(from = 0, to = 1, by = 0.1)
seq(0, 10, length.out = 5)  # 5 evenly spaced values

# Repeating values
rep(5, times = 3)
rep(c(1, 2), each = 3)
rep(c(1, 2), times = 3)
```

---

## Basic Functions

Functions are operations that take inputs (arguments) and return outputs.

```{r}
# Calculate the mean (average)
mean(c(10, 20, 30, 40, 50))

# Calculate the sum
sum(1:100)

# Find the minimum and maximum
min(c(5, 3, 8, 1))
max(c(5, 3, 8, 1))

# Round numbers
round(3.14159, digits = 2)
floor(3.7)  # Round down
ceiling(3.2)  # Round up
```

---

## Function Arguments

Functions can have multiple arguments to control their behavior.

```{r}
# Named arguments
mean(x = c(1, 2, 3), na.rm = TRUE)

# Positional arguments (order matters)
round(3.14159, 2)  # Same as round(x = 3.14159, digits = 2)

# Default arguments
# If you don't specify digits, round() uses digits = 0
round(3.14159)
```

---

## Missing Values (NA)

R uses `NA` to represent missing values.

```{r}
# Vector with missing values
data <- c(10, NA, 20, 30, NA)
data

# Many functions have an na.rm argument to remove NAs
mean(data)               # Returns NA
mean(data, na.rm = TRUE) # Calculates mean excluding NAs

# Check for missing values
is.na(data)
sum(is.na(data))  # Count missing values
```

---

## Getting Information About Objects

```{r}
# Create a vector
numbers <- c(5, 10, 15, 20, 25)

# Get information about the object
length(numbers)  # How many elements
class(numbers)   # What type of object
str(numbers)     # Structure of the object
summary(numbers) # Summary statistics
```

---

## Writing Comments

Comments help document your code and explain what it's doing.

```{r}
# This is a comment - it won't be executed
x <- 10  # You can put comments at the end of lines too

# Multi-line comments for longer explanations
# like this one that continues
# across several lines

2 + 2  # This adds two numbers
```

<div class="tip">
<strong>Tip:</strong> Well-commented code is easier to understand, debug, and maintain.
</div>

---

## Exercise: Basic Commands

Try running these commands in RStudio:

1. Create variables `height` and `width` with values 7 and 5.
2. Calculate the area of a rectangle using these variables.
3. Create a vector of the first 10 even numbers.
4. Calculate the mean, min, and max of this vector.
5. Create a logical vector testing which numbers in the vector are greater than 10.

---

# Questions? {data-background="#3a0ca3"}

### Next topic: Using Excel and CSV Files

---

## Additional Resources

- [R for Data Science: Basic R](https://r4ds.had.co.nz/workflow-basics.html)
- [Quick-R: Basic Operations](https://www.statmethods.net/management/operators.html)
- [DataCamp: Introduction to R](https://www.datacamp.com/courses/free-introduction-to-r)
- [RStudio Primers](https://rstudio.cloud/learn/primers)

```{js, echo=FALSE}
// Print PDF setup
function printPDF() {
  let link = document.createElement('link');
  link.rel = 'stylesheet';
  link.type = 'text/css';
  link.href = window.location.search.match(/print-pdf/gi) ? 'reveal.js/css/print/pdf.css' : 'reveal.js/css/print/paper.css';
  document.getElementsByTagName('head')[0].appendChild(link);
}
printPDF();
```
